\section{Reproducibility Statement}
\label{sec:reproducibility}

RPL-ClusterIDS follows reproducible-research practices aligned with Computer Networks expectations.

\subsection{Artifact Availability}
The following artifacts are maintained in the public repository:
\begin{itemize}
  \item Repository: \url{https://github.com/madani-belacel/IDS-IOT}
  \item Release v1.0 (complete ZIP package): \url{https://github.com/madani-belacel/IDS-IOT/releases/tag/v1.0}
  \item Contiki-NG firmware: \path{code_source_RPL_ClusterIDS/} (variants B1, B2, B3, CLUSTERIDS, and ablation variants);
  \item Campaign specification: scenarios, attacks, seeds, metric definitions;
  \item Data layout: \path{data/estimated/} (synthetic templates for pipeline testing), \path{data/real/} (pilot campaign parsed outputs);
  \item Figure catalog: \path{figures_manifest.csv} with per-figure provenance.
\end{itemize}
All code, simulation scenarios, raw and parsed logs, and figure-generation scripts are available in the release v1.0. A persistent DOI is planned for the full campaign archive and will be linked once the artifact is deposited.

\subsection{Build Instructions}
The submission manuscript is compiled from \texttt{main.tex}:
\begin{verbatim}
pdflatex main && bibtex main &&
pdflatex main && pdflatex main
\end{verbatim}
Simulation execution requires Ubuntu with Contiki-NG and Cooja (Phase~2).
The launcher \path{SIMULATION_CAMPAIGN_READY/run_campaign.sh} orchestrates the full factorial design in \path{campaign_matrix.tsv}.

\subsection{From Logs to Figures}
The pipeline is:
\begin{enumerate}
  \item Cooja serial logs $\rightarrow$ \path{code_source_RPL_ClusterIDS/simulations/scripts/parse_cooja_ids_metrics.py};
  \item Aggregated CSV $\rightarrow$ \path{scripts/statistics/compute_statistics.py};
  \item Statistics $\rightarrow$ \path{data/real/parsed/agg/aggregate_figures.sh} $\rightarrow$ Figs.~4--11;
  \item Provenance documented in \path{sim/DATA_PROVENANCE.md}.
\end{enumerate}

\subsection{Result Status Convention}
Results come from the pilot Cooja campaign (50-node grid, 21 seeds for the CLUSTERIDS variant, 3 seeds for the ablation study; five attack scenarios).
Parsed logs are archived under \path{data/real/parsed/}. CPU, RAM, and energy values in the pilot campaign are preliminary estimates derived from the simulation pipeline; real Energest measurements will replace them in the full campaign. Detection rates and FPR reported here reflect the simulated IDS overlay state captured by the Cooja tick-based metrics pipeline and should be treated as preliminary pilot measurements.

\subsection{Data Availability}
All firmware, campaign scripts, parsed datasets, and figure-generation code are publicly available at \url{https://github.com/madani-belacel/IDS-IOT/releases/tag/v1.0}.
\path{data/real/parsed/} contains the parsed CSV outputs from the pilot campaign.
